Examining a Quarter Century of Publishing Trends in Social Work Research:

### 2015-05-27 manuscript.Rmd
Updated methods section.

### Extraction of article records

Our list of social work journals was guided by the work of Hodge and Lacasse
(2011).  More specifically, Hodge and Lacasse (2011) identified 84 disciplinary
social work journals based on a variety of sources, including _An Author’s
Guide to Social Work Journals_ (NASW Press, 1997), Thyer’s (2005) more recent
listing of social work periodicals, and Genamics JournalSeek
(http://journalseek.net/). Hodge and Lacasse (2011) examined the mission and
aims of each journal, and eliminated journals that were specific to another
field or had an had an inter-disciplinary focus.  For the current study, our
primary search query included all journal titles in this list.

To help ensure comprehensive coverage of all possible core social work
journals, we created a supplemental search query that extracted journal titles
with any of the following terms:

@@

Using this search queries, we extract journal article recrods from three major
databases on the EbscoHost platform:  PsycINFO, Social Service Abstracts, and
Social Work Abstracts.  Because of known gaps in indexing in databases (Holden,
Barker, Kuppens, Rosenberg & LeBreton, 2014; Holden, Barker, Covert-Vail,
Rosenberg, & Cohen, 2009), we also searched ProQuest, which linked to an
additional 46 minor databases, including (but not limited to): ERIC,
Sociological Abstracts and Worldwide Political Science Abstracts.  In every
search, we used a filter in attempt to extract only article records that were
classified as _journal articles_, thereby excluding other forms of scholarly
communication that was not relevant to the current study (e.g., book reviews,
editorials, obituaries, etc.).  We limited the timeframe to a 25 year period,
from 1989 to 2013 (inclusive).  Articles published in years 2014 and 2015 were
excluded because of delays in indexing.  Search results were exported in
batches of article records based on the restrictions of the platform.  Article
records were text files in a _generic bibliographic format_.  The article
records contained various meta-data based on the database from which the
article record was extracted.   Meta-data include (but are not limited to):
article title, journal title, publication year, author name(s), author
affiliation(s), abstract, keywords, methodological classification, funding
source, location of study, subject groups, digital object identifier (DOI),
number of references, number of pages, etc.  These files were post-processed
into a structured database using a set of scripts written in the R statistical
programming language.  The initial search resulted in `r
pretty(nrow(article.count.initial))` article records from `r
journal.count.initial`.

The search queries were purposefully specified to be overly inclusive to ensure
full coverage of all possible social work journals contained in the major and
minor databases.  Additionally, a visual inspection of the article records
revealed problems in the original indexing of articles that requires subsequent
data cleaning.  Thus, our data cleaning procedures focused separately on
journal titles and article records.

### Data cleaning: Journal titles

The first step of data cleaning involved fixing journal title names due to
discrepant errors in indexing. For example, _Journal of Gay and Lesbian Social
Services_ was indexed as a journal separate from _Journal of Gay & Lesbian
Social Services_ (use of "and" vs. "&").  Other examples involved journals
indexed with and without subtitles, or journals with and without the word "The"
at the beginning of the title.  Additionally, some journals changed titles over
their history.  For example, _Journal of Technnology in Human Services_ is
formerly known as _Computers in Human Services_.  In these situations, the
former titles were merged with the current titles.

Many journals from allied health disciplines used one of the supplemental
search terms in special editions, which were also included in the journal
title.  Thus, many non-social work journals were captured in the extraction
process.  To resolve this issue, we created a list of all journal titles that
were not part of the core list defined by Hodge and Lacasse (2011).  Study
authors reviewed these journal titles and discussed whether each candidate
title should be retained or excluded.  When disagreements occurred, the study
authors reviewed the mission and aims of the journals, names of editorial board
members, and focus of the articles.  A consensus was reached on all journal
titles to be excluded and retained.   After these procedures, the number of
article records and journal titles was reduced to `r
pretty(nrow(article.count.journalsCleaned))` and `r
journal.count.journalsCleaned`.

### Data cleaning:  Article records

Our search procedures involved the use of three major databased and 46 minor
databases, which resulted in some duplication of article records.  Thus, a
matching algorithm was constructed to identify and remove all duplicate article
records. As previously noted, the filtering of only journal articles was not
successful due to errors in the original indexing.  Thus, other scholarly
communications were captured in the extractio process.  These other scholarly
communications were readily identified due to specific patterns in the titles,
such as the terms "Book Review," "From the Editor," and "Obituary."  Given the
number of article records in the database, it was not feasbile to manually
extract these recrods.  Instead, we created a separate database with
representative examples of article records that we wanted to extract.  We also
added representative examples of article records to be retained.  We then wrote
a series of _regular expressions_ to extract the problematic article records.
Regular expressions are a sequence of characters that form search patterns,
allowing extraction of records that match the specified pattern.  In other
words, it is a more sophisticated implementation of the search function common
to all major word processors.  We tested our regular expressions on the test
database, achieving > 95% accuracy with regard to retaining and extracting
article records.  These regular expressions were then applied to the full
database.  We then excluded article records if they were missing essential
meta-data, including author name(s), journal title, article title, and
publication field.  As an additional requirement, we required all articles to
be $\geq 3$ pages in length.   Finally, we eliminated all social work journals
with total article counts of < 10 for the entire 25 year window (_Issues in
Social Work Education_, _Maatskaplike Werk/Social Work_, _Pediatric Social
Work_, and _Critical Social Work_).  These cleaning procedures resulted in a
final database of `r pretty(articles.final)` articles and `r journals.final`
journals.

### Data quality checks

We performed a variety of checks on the final database.  Because the size of
the database, it was not feasible to manually inspect all records.  Thus, we
randomly sampled article records from the database and cross-checked the data
we procured with the information listed with the journal's homepage.  We also
inspected a scatter plot of the number of article records for each journal by
the number of years the journal appeared within the 25 window timeframe.
Journals that exhibited significant deviation from the regression line were
inspected to ensure that our cleaning procedures did not systematically exclude
article records or retain other scholarly communications.  All observed
discrepancies were verified as differences in journals actual publication
output.  Finally, we also made checks of number of journal article records for
various journals indexed in the _Web of Science_.  Although the counts were
close -- i.e., within 10% -- similar indexing problems observed from our record
extraction were also observed in the _Web of Science_ database, such as
misclassification of book reviews as journal articles.  Thus, we are unable to
quantify the actual amount of error in our database is due to problems in
indexing.  At the same time, we are relying on the same source of data that
social work researchers use to inform their work.  Too much cleaning and
post-processing of the search results can give rise to validity issues because
social work researchers are using an information source that contain the errors
we have attempted to eliminate.  The issue of reliability and validity are
given further attention in the discussion section of this study.

### Analyses

>>>>

## Social work journal titles

The primary unit of analysis for this study was bibliographic records of social
work journal articles (hereafter referred to as ``article records``).  The
first step in the extraction process involved constructing a comprehensive list
of social work journal titles.  We used the work of Hodge and Lacasse (2011) as
our base list.  Hodge and Lacasse (2011) compiled a list of 84 unique social
work journal titles using a variety of sources, includingt the book _An
Author’s Guide to Social Work Journals_ (NASW Press, 1997), Thyer’s (2005) more
recent listing of social work periodicals, and the website _Genamics
JournalSeek_ (http://journalseek.net/). Hodge and Lacasse (2011) examined the
mission and aims of each journal, and eliminated journals that were specific to
another field or had an inter-disciplinary focus.  All social work journals
identified by Hodge and Lacasse (2011) were included on our base list.

To ensure comprehensive coverage, we also conducted a search of the Internet
and other bibliographic databases to identify other journals that may be
considered a disciplinary social work journal. The search terms included:

@@

Journal titles that were identified in the supplemental search were reviewed
by the study team. When disagreements occurred, the authors reviewed the
mission and aims of the journals, names of editorial board members, and content
of representative articles.  A consensus was reached on all journal titles to
be excluded and retained.  In this process, an additional XX journal were
identified, producing a final target list of XX journals.

## Extraction of article records

After the target list of social work journal titles was finalized, we attempted
to identify all article records from these journals using three major databases
on the EbscoHost platform (PscyINFO, Social Science Abstracts, and Social Work
Abstracts) and 46 minor databases on Proquest (e.g., ERIC, Sociological
Abstracts, Worldwide Political Science Abstracts).  Using the ``publication
source`` fields of the search interface, we searched for article records that
contained the names of each social work journal.  Searches were limited to a
quarter century time-frame, 1989 to 2013.  We excluded years 2014 and 2015 due
to delays in indexing of records that were observed.  In the extraction
process, we also limited the records to only those classified as "journal
articles," thereby excluding other types of scholarly communications, such as
editorials, comments on articles, letters to the editor, book reviews, and
obituaries.

Search results were saved and exported as plain text files in _generic
bibliographic format_. The article records contained meta-data, including (but
not limited to): article title, journal title, publication year, author
name(s), author affiliation(s), abstract, keywords, methodological
classification, funding source, location of study, subject groups, digital
object identifier (DOI), number of references, number of pages, etc. It should
be noted that the databases provided different pieces of meta-data, or
meta-data in varying formats.  Thus, extensive post-process of the article
records was necessary to make the article records comparable and analyzable.

## Post-processing of article records

Articles were post-processed from the _generic bibliographic format_ to a
structured database with computer scripts authored specifically for this study
using the programming language R. Post-processing of the files involved
addressing numerous problems that occurred during the original indexing of the
article records.  For example, journal titles that included the word ``and''
sometimes used an amersand (&). Other examples included inconsistent use of the
word ``The'' at the beginning of the title, or the inconsistent use of
subtitles.  It should be noted that some journals changed titles over their
publication history.  For example,  _Journal of Technology in Human Services_
is formerly known as _Computers in Human Services_.  All historical titles were
merged with their current titles and analyzed as a single journal in this
study. The source code provides complete details of all the procedures to
resolve discrepancies and historical titles.

In addition to the foregoing discrepancies, we also found other problems among
the extracted set of article records.  For example, the filtering of article
records did not successfully exclude all other types of documents (e.g.,
editorials, book reviews, and obituaries).  Thus, we used a set of text
matching procedures called _regular expressions_ to identify other documents
that were incorrectly classified as journal articles.  We observed > 95%
accuracy identifying non-article records using a separate database designed
specifically for script testing.  We excluded article records if they were
missing essential meta-data for this study, including author name(s), journal
title, article title, and year of publication.  Finally, we eliminated all
social work journals with total article counts of < 10 for the entire 25 year
window (_Issues in Social Work Education_, _Pediatric Social Work_, and
_Critical Social Work_).

## Data quality checks

We performed a variety of checks on the final database.  Because the size of
the final database, it was not feasible to manually inspect all records.  Thus,
we randomly sampled article records from the database and cross-checked journal
meta-data with information listed on the journal's web page.  To help build
confidence in the comprehensiveness of the constructed database, we obtained
aggregate publication data from Scopus and made systematic comparisons with the
database constructed for this project.  Scopus data was selected because it is
the single largest abstract and citation database of scientific journals and
was not linked to either the EbscoHost or ProQuest databases in the record
extraction process.  The systematic comparisons are

In addition to post-processing original article records, R was also used for
all other data analysis.  R was an ideal analytic tool because it is
open-source and has a suite of related tools for producing reproducible
research.  More specifically, all the source code is written using Rmarkdown --
an open-source system that integrates data management, data analysis, and the
written report.  The source code and data extracts is also curated at openICPSR
of the Inter-university Consotrium for Political and Social Research of the
University of Michigan.  The source code is also available from the first
author's GitHub repository, where the code will be maintained and updated to
support future research projects. It should be noted that the full data files
cannot be made publicly available because these are considered proprietary data
of the database services.  However, to ensure full compliance with the best
practices of reproducible research (cite King), we have [BRYAN -- DESCRIBE HERE
HOW THE PS JOURNAL HANDLES THIS ISSUE ... ASSUME THE EDITOR IS ON-BOARD WITH
THIS APPROACH].
